---
title: "P-splines"
author: "Monica Alexander"
date: "22/03/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

In this lab you'll be fitting a second-order P-Splines regression model to foster care entries by state in the US, projecting out to 2030. 

```{r}
library(tidyverse)
library(here)
library(rstan)
library(tidybayes)
source(here("code/getsplines.R"))
library(geofacet)
library(ggplot2)
```



Here's the data

```{r}
d <- read_csv(here("data/fc_entries.txt"))
```

## Question 1

Make a plot highlighting trends over time by state. Might be a good opportunity to use `geofacet`. Describe what you see in a couple of sentences. 




```{r}
ggplot(d, aes(year, ent, fill = ent)) +
  geom_line() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(strip.text.x = element_text(size = 5)) +
  labs(title = "foster care entries per capita trends over time by state",
       x = "years", y = "entries per capita")+
  facet_geo(~ state, grid = "us_state_grid2",scales = "free_y")
```

We see the in many states there's an increasing trend of foster care entries. I suspect that there's a correlation between adjacent states.


## Question 2

Fit a hierarchical second-order P-Splines regression model to estimate the (logged) entries per capita over the period 2010-2017. The model you want to fit is

$$
\begin{aligned}
y_{st} &\sim N(\log \lambda_{st}, \sigma^2_{y,s})\\
\log \lambda_{st} &= \alpha_kB_k(t)\\
\Delta^2\alpha_k &\sim N(0, \sigma^2_{\alpha,s})\\
\log \sigma_{\alpha,s} &\sim N(\mu_{\sigma}, \tau^2)
\end{aligned}
$$
Where $y_{s,t}$ is the logged entries per capita for state $s$ in year $t$. Use cubic splines that have knots 2.5 years apart and are a constant shape at the boundaries. Put standard normal priors on standard deviations and hyperparameters. 

```{r}
x.i <- d[1:8,]$ent
I <- 2.5 # between-knot length
res <- getsplines(x.i, I = I) # a function from distortr, to get splines of constant shape
B.ik <- res$B.ik
x.i[1] # look at first value of BMI

```
```{r}
I <- 2.5
res <- getsplines(d$year, I = I)
stan_data <- list(nb = ncol(res$B.ik),
                   ns = length(unique(d$state)),
                   ny = length(unique(d$year)),
                   Y = matrix(log(d$ent_pc), ncol = 51),
                   B = res$B.ik[1:length(unique(d$year)), ])
mod <- stan(data = stan_data,
             file = "spline.stan",
             iter = 5000,
             seed = 2)
```



## Question 3

Project forward entries per capita to 2030. Pick 4 states and plot the results (with 95% CIs). Note the code to do this in R is in the lecture slides. 

```{r}
years <- 2010:2017
proj_years <- 2018:2030
B.ik_full <- getsplines(c(2010:2017, proj_years), I = 2.5)$B.ik
K <- ncol(res$B.ik)
K_full <- ncol(B.ik_full)
proj_steps <- K_full - K
alphas <- extract(mod)[["alpha"]]
sigmas <- exp(extract(mod)[["log_sigma_alpha"]])
sigma_ys <- extract(mod)[["sigma"]]
nsims <- nrow(alphas)
states <- unique(d$state)
alphas_proj <- array(NA, c(nsims, proj_steps, length(states)))
set.seed(1098)
for(j in 1:length(states)){
  first_next_alpha <- rnorm(n = nsims, mean = 2*alphas[, K,j] - alphas[, K-1,j], sd = 
                              sigmas[,j])
  second_next_alpha <- rnorm(n = nsims, mean = 2*first_next_alpha - alphas[, K,j], sd =
                               sigmas[,j])
alphas_proj[,1,j] <- first_next_alpha
alphas_proj[,2,j] <- second_next_alpha
for(i in 3:proj_steps){

  alphas_proj[,i,j] <- rnorm(n = nsims,
                             mean = 2*alphas_proj[,i-1,j] - alphas_proj[,i-2,j],
                             sd = sigmas[,j])
  }
}
y_proj <- array(NA, c(nsims, length(proj_years), length(states)))
for(i in 1:length(proj_years)){
  for(j in 1:length(states)){
    all_alphas <- cbind(alphas[,,j], alphas_proj[,,j] )
    this_lambda <- all_alphas %*% as.matrix(B.ik_full[length(years)+i, ])
    y_proj[,i,j] <- rnorm(n = nsims, mean = this_lambda, sd = sigma_ys[,j])
    }
}


res1 <- mod %>%
  gather_draws(Y_rep[condition1, condition2]) %>%
  median_qi()
colnames(res1)[1:2] <- c("year", "state")
index <- which(states %in% c("California", "Mississippi", "Ohio", "Texas"))
res2 <- matrix(NA, nrow = 663, ncol = 5)
res2[, 1] <- rep(1:length(proj_years), each = 51)
res2[, 2] <- rep(1:51, length(proj_years))
colnames(res2) <- c("year", "state", ".value", ".lower", ".upper")
res2 <- data.frame(res2)
for (i in 1:663) {
  res2[i, ".value"] <- median(y_proj[, res2$year[i], res2$state[i]])
  res2[i, ".lower"] <- quantile(y_proj[, res2$year[i], res2$state[i]], 0.025)
  res2[i, ".upper"] <- quantile(y_proj[, res2$year[i], res2$state[i]], 0.975)
}
res1_plot <- res1 %>%
  filter(state %in% index)
res1_plot$year <- rep(years, each = 4)
res1_plot$state <- rep(c("California", "Mississippi", "Ohio", "Texas"),
                       length(unique(res1_plot$year)))
res2_plot <- res2 %>%
  filter(state %in% index)
res2_plot$year <- rep(proj_years, each = 4)
res2_plot$state <- rep(c("California", "Mississippi", "Ohio", "Texas"),
                       length(unique(res2_plot$year)))
d_plot <- d %>%
  filter(state %in% c("California", "Mississippi", "Ohio", "Texas"))
d_plot$log_ent_pc <- log(d_plot$ent_pc)
ggplot(d_plot, aes(year, log_ent_pc)) +
  geom_point(aes(color = state)) +
  geom_line(data = res1_plot, aes(year, .value, col = state)) +
  geom_ribbon(data = res1_plot, aes(x = year, y = .value, ymin = .lower, ymax = .upper, fill =
                                      state), alpha = 0.2) +
  theme_bw() +
  geom_line(data = res2_plot, aes(year, .value, col=state)) +
  geom_point(data = res2_plot, aes(year, .value, col=state)) +
  geom_ribbon(data = res2_plot, aes(y = .value, ymin = .lower, ymax = .upper, fill=state), alpha
              = 0.2) +
  theme_bw() +
  labs(title = "Estimated and projected entries per capita second order P splines",
       y = "log entries per capita")
```



